% !TeX root = ../main.tex
\section{Type inference}
%   - Juicios de tipado $\Gamma \vdash e : \tau$
%   - Reglas de inferencia para expresiones y comandos
%   - Algoritmo W (Hindley-Milner) si aplica
%   - Ejemplos sobre la gramática del capítulo anterior
Type inference is the process by which a compiler deduces the type of an expression without explicit annotations from the programmer. the classical algorithm for the simply typed lambda calculus (and its extensions) is \emph{Hindley Milner} (HM), which works by:

\begin{enumerate}
  \item Assigning a fresh type variable $\alpha_i$ to each sub-expression
        whose type is unknown.
  \item Generating \emph{constraints} of the form $\tau_1 = \tau_2$ from the
        typing rules.
  \item Solving the constraints by \emph{unification}.
\end{enumerate}

if unification fails, the program contains a type error; if it succeeds, the resulting substitution gives the most general type.

\subsection{Typing rules for references}
\[
\infer[\textsc{T-Deref}]
  {\Gamma \vdash \deref{l} : T}
  {\Gamma \vdash l : \tref{T}}
\qquad
\infer[\textsc{T-Assign}]
  {\Gamma \vdash \assign{l}{e} : \unit}
  {\Gamma \vdash l : \tref{T} \quad \Gamma \vdash e : T}
\qquad
\infer[\textsc{T-While}]
  {\Gamma \vdash \whiledo{B}{C} : \unit}
  {\Gamma \vdash B : \tybool \quad \Gamma \vdash C : \unit}
\]
\subsection{solved example}
$\lambda f.\, f\, f$ does not typecheck

Consider \lstinline|proc(f) zero?((f f))|.  We try to infer the type of $f$.
 
\begin{itemize}
  \item Let $f : \alpha$.
  \item The expression $f\, f$ requires $f$ to be a function type, so
        $\alpha = \alpha \to \beta$ for some fresh $\beta$.
  \item But $\alpha = \alpha \to \beta$ has no finite solution: expanding
        repeatedly yields $\alpha = (\alpha \to \beta) \to \beta =
        ((\alpha\to\beta)\to\beta)\to\beta = \cdots$
\end{itemize}

unification fails (occurs check). Therefore the expression has no type in the simply typed lambda calculus without recursive types.

\subsection{Scala generics}
\begin{lstlisting}
case class MyPair[A,B](x:A,y:B)
val p = MyPair(1, "Scala")
def id[1](x:T)=x
val q = id(1)
\end{lstlisting}


the definition of \lstinline|id| introduces a \texttt{polymorphic} type: it can be applied to any type , and each call site independently instantiates a $T$.

\begin{lstlisting}
B  = !n < 5
B1 = !term <= 2
C1 = res := !n
C2 = c:=!a+!b; res:=!n; !c; a:=!b; b:=!c
C  = if B1 then C1 else C2; n:=!n+1
a:=1; b:=1; while B do C
\end{lstlisting}

reading off types from the operations performed:
\begin{itemize}
  \item $n, \mathit{res}, a, b, c : \tref{\tyint}$ (assigned integers and
        dereferenced in arithmetic).
  \item $\mathit{term} : \tref{\tyint}$ (compared with $2$).
  \item $B, B_1 : \tybool$.
  \item $C_1, C_2, C : \unit$ (commands produce no value).
  \item The whole program : $\unit$.
\end{itemize}

\subsection{Exercises}
 
Infer the type of the S combinator:
$\lambda x.\,\lambda y.\,\lambda z.\; x\,z\,(y\,z)$.
What constraints does unification generate?
 
Consider $\lambda f.\,\lambda x.\; f\,(f\,x)$.  Does it typecheck?  If so,
give its most general type.
 
What types are inferred for \lstinline|f|, \lstinline|result|, and
\lstinline|pairs| in the following Scala snippet?
\begin{lstlisting}
def f[A](lst: List[A]) = lst.head
val result = f(List(1,2,3))
val pairs  = List((1,"a"), (2,"b"))
\end{lstlisting}

Given $x : \tref{\tyint}$ and $b : \tref{\tybool}$, does the following
expression typecheck?  If so, give its type.
\[
  \assign{b}{\deref{x}=0}\;;\;\ift{\deref{b}}{\assign{x}{1}}{\mathbf{skip}}
\]





