% !TeX root = ../main.tex
\section{Type safety}
% Sugerencias de contenido:
%   - Enunciado de progreso (\emph{progress}):
%     si $\vdash e : \tau$ entonces $e$ es un valor o existe $e'$ con $e \reduces e'$.
%   - Enunciado de preservación (\emph{preservation}):
%     si $\vdash e : \tau$ y $e \reduces e'$ entonces $\vdash e' : \tau$.
%   - Lema de canonicidad de valores
%   - Lema de sustitución
%   - Esquema de la demostración por inducción estructural
Type safety is the formal guarantee that well-typed programs do not (go wrong), they never reach a state that is undefined or meaningless. It is established through two theorems that work together: \textbf{progress} and \textbf{preservation}.

\subsection{progress}
A well-typed expression is never \emph{stuck}, it is either already a value, or it can take a step.
\begin{theorem}[progress]
If $\vdash e : \tau$ and $e \reduces e'$, then $\vdash e' : \tau$.
\end{theorem}

Intuitively, this rules out situations like applying an integer as if it were a function, a well-typed program always has somewhere to go.

\subsection{preservation}
Taking a step does not change the type of an expression.

\begin{theorem}[preservation]
If $\vdash e : \tau$ and $e \reduces e'$, then $\vdash e' : \tau$
\end{theorem}

Together, progress and preservation give us the \textbf{type safety} property: evaluation of a well-typed expression either produces a value of the expected type, or diverges; but it never crashes.

\subsection{Supporting lemmas}
The proofs of the both theorems rely on two key lemmas.
\begin{lemma}[Canonicity]
If $\vdash v : \tau$ and $v$ is a value, then the \emph{shape} of $v$
is determined by $\tau$. For example:
\begin{itemize}
    \item If $\tau = \mathit{Bool}$, then $v$ is either $\mathit{true}$
    or $\mathit{false}$.
    \item If $\tau = \tau_1 \to \tau_2$, then $v = \lambda x.\, e$ for
    some $x$ and $e$.
\end{itemize}
\end{lemma}

This lemma is what lets us case-split on the shape of a value during a proof, we know exactly what forms are possible.

\begin{lemma}[Substitution]
If $\Gamma, x : \tau' \vdash e : \tau$ and $\vdash v : \tau'$, then
$\Gamma \vdash e[x := v] : \tau$.
\end{lemma}

in plain terms, substituing a well-typed value for a variable preserves typing. This is the key step in proving Preservation for $\beta$ reduction.

\subsection{proof strategy: structural induction}
Both theorems are proved by \textbf{structural induction} on the typing derivation. The general scheme is:

\begin{enumerate}
    \item \textbf{Base cases:} handle values and variables directly, these are straightforward since no reduction is possible.
    \item \textbf{Inductive cases:} for each typing rule (application,
    abstraction, conditionals, etc.), assume the result holds for the
    subexpressions (\emph{induction hypothesis}), then show it holds for
    the whole expression.
    \item \textbf{Use the lemmas:} Canonicity tells you what shape a
    value has; Substitution handles the $\beta$-reduction case in
    Preservation.
\end{enumerate}

the structure of the proof closely mirrors the structure of the language, one case per typing rule. Once you see the patternm it becomes mechanical.
