% !TeX root = ../main.tex
\section{Axiomatic semantics}
% Sugerencias de contenido:
%   - Tripletas de Hoare $\{P\}\, C\, \{Q\}$
%   - Reglas: asignación, secuencia, condicional, while, consecuencia
%   - Invariantes de bucle
%   - Corrección parcial vs total
%   - Precondición más débil (weakest precondition, $\mathit{wp}$)
%   - Ejemplos: factorial, suma de un arreglo
Axiomatic semantics takes a different approach from the other two styles: instead of describing how a program executes or what function it computes, it gives meaning to programs through \textbf{logical assertions} about their behavior. The central idea is to reason about what is true \emph{before} and \emph{after} a program runs.

\subsection{Hoare triples}
The basic unit is the \textbf{Hoare triple}:
\[
  \{P\}\; C\; \{Q\}
\]
where $P$ is the \textbf{precondition}, $C$ is a command, and $Q$ is the \textbf{postcondition}. Read it as: ``if $P$ holds before running $C$, and $C$ terminates, then $Q$ holds afterward.''

For example:
\[
  \{x = 5\}\; x := x + 1\; \{x = 6\}
\]

\subsection{inference rules}
The power of hoare logic comes from its \textbf{compositional rules}, one per language construct.

\begin{itemize}
  \item \textbf{Assignment:}
  \[
    \{P[x := e]\}\; x := e\; \{P\}
  \]
  Work backwards: to have $P$ after the assignment, you need $P$ with
  $e$ substituted for $x$ before it.

  \item \textbf{Sequence:}
  \[
    \frac{\{P\}\; C_1\; \{R\} \quad \{R\}\; C_2\; \{Q\}}
         {\{P\}\; C_1;\, C_2\; \{Q\}}
  \]
  The postcondition of the first command becomes the precondition of
  the second.

  \item \textbf{Conditional:}
  \[
    \frac{\{P \wedge b\}\; C_1\; \{Q\} \quad \{P \wedge \neg b\}\;
    C_2\; \{Q\}}
         {\{P\}\; \texttt{if}\; b\; \texttt{then}\; C_1\;
         \texttt{else}\; C_2\; \{Q\}}
  \]
  Each branch gets the precondition strengthened with the condition or
  its negation.

  \item \textbf{While:}
  \[
    \frac{\{I \wedge b\}\; C\; \{I\}}
         {\{I\}\; \texttt{while}\; b\; \texttt{do}\; C\;
         \{I \wedge \neg b\}}
  \]
  The loop \textbf{invariant} $I$ must hold before the loop, be
  preserved by every iteration, and together with the negated condition
  gives the postcondition when the loop exits.

  \item \textbf{Consequence:}
  \[
    \frac{P \Rightarrow P' \quad \{P'\}\; C\; \{Q'\} \quad
    Q' \Rightarrow Q}
         {\{P\}\; C\; \{Q\}}
  \]
  Allows strengthening preconditions and weakening postconditions, essential for connecting rules together in a proof.
\end{itemize}


\subsection{loop invariants}
Finding a good loop invariant is often the hardest part od a hoare logic proof. A valid invariant $I$ must satisfy three conditions:
\begin{enumerate}
  \item $I$ holds \textbf{before} the loop begins.
  \item If $I$ holds and the condition $b$ is true, then after one
  iteration of the body, $I$ still holds.
  \item When the loop exits ($b$ is false), $I \wedge \neg b$ implies
  the desired postcondition.
\end{enumerate}

\subsection{partial vs.\ total correctness}
There are two flavors of correctness:
\begin{itemize}
  \item \textbf{Partial correctness} $\{P\}\, C\, \{Q\}$: if $C$
  terminates, then $Q$ holds. It says nothing about \emph{whether}
  $C$ terminates.
  \item \textbf{Total correctness} $[P]\, C\, [Q]$: $C$ is guaranteed
  to terminate \emph{and} $Q$ holds afterward.
\end{itemize}

total correctness requieres an additional argument, typically a \textbf{variant}, a quantity that strictly decreases with each iteration and is bounded below.

\subsection{weakness preconditions}

Rather tahn guessing a precondition, we can \emph{compute} the weakest one that makes a triple valid. The \textbf{weakest precondition} $\mathit{wp}(C, Q)$ is the least restrictive $P$ such that $\{P\}\, C\, \{Q\}$ holds.

The key cases are:
\begin{align*}
  \mathit{wp}(x := e,\; Q) &= Q[x := e] \\
  \mathit{wp}(C_1;\, C_2,\; Q) &= \mathit{wp}(C_1,\;
  \mathit{wp}(C_2, Q)) \\
  \mathit{wp}(\texttt{if}\; b\; \texttt{then}\; C_1\;
  \texttt{else}\; C_2,\; Q)
    &= (b \Rightarrow \mathit{wp}(C_1, Q))
       \wedge (\neg b \Rightarrow \mathit{wp}(C_2, Q))
\end{align*}
For \texttt{while} loops, $\mathit{wp}$ cannot be computed
mechanically in general, this is where loop invariants come back in.

\subsection{Examples}

\subsubsection{Factorial}

Consider a program that computes $n!$ iteratively, storing the result
in $r$. A suitable invariant is:
\[
  I \;:\; r \cdot i! = n! \;\wedge\; 0 \leq i \leq n
\]
Before the loop: $r = 1$, $i = n$, so $1 \cdot n! = n!$ — the
invariant holds. Each iteration multiplies $r$ by $i$ and decrements
$i$, preserving $r \cdot i! = n!$. When $i = 0$, we get $r = n!$.


\subsubsection{Array Sum}

For a program that sums the elements of an array $a[0..n-1]$ into
variable $s$:
\[
  I \;:\; s = \sum_{k=0}^{i-1} a[k] \;\wedge\; 0 \leq i \leq n
\]
The invariant tracks a partial sum. When the loop exits with $i = n$,
we have $s = \sum_{k=0}^{n-1} a[k]$, which is exactly the full sum.





